Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2022-08-16 Presentations/YESconference.tex

\begin{frame}{Big Picture}
\begin{itemize}
    \item Longstanding topic in economics: relationship between technology and employment
    \vspace{2mm}
    \begin{itemize}
        \item Displacement?
        \item Productivity?
    \end{itemize}
    \vspace{3mm}
    \item Investigated in many settings
    \vspace{3mm}
    \item The answer depends heavily on the setting
    \vspace{3mm}
    \item This paper: specific healthcare setting
\end{itemize}

\end{frame}

\begin{frame}[fragile]{What is an Electronic Health Record?}
``An electronic health record is the systematized collection of patient and population electronically stored health information in a digital format"

@@

What it can do:
\begin{itemize}
    \item Paper $\rightarrow$ computerization
    \item Communication
    \item Decision making assistance
\end{itemize}

\end{frame}

@@

Significant policy push for EHR implementation: HITECH Act, 2008 provided financial incentive for hospitals to implement EHRs
\begin{itemize}
    \item \textcolor{blue}{The percentage of hospitals with basic EHR capability rose from 9$\%$ in 2008 to 84$\%$ in 2015.} \scriptsize (Health IT Dashboard)
\end{itemize}

\end{frame}

@@

    \includegraphics[scale=.3]{graphics/News Clip2.PNG}
\end{center}

\end{frame}

\begin{frame}{This Paper}

Did EHR implementation in hospitals affect physician labor market decisions?
\begin{itemize}
    \item Intensive Margin / Productivity
    \item Extensive Margin / Retirement
    \item Hospital $\rightarrow$ Office
\end{itemize}

\end{frame}

\section{Contribution}

\begin{frame}{Literature}
\small
\begin{columns}
\setlength{\tabcolsep}{-5pt}
    \column{0.4\textwidth}
        \centering
        \underline{ Technology $\rightarrow$ Labor }
        \vspace{-1mm}
        \begin{itemize}
            \item Erosion effect vs. wage effect \\ \vspace{1mm}
            \tiny (Zeira $\&$ Joseph 2011, EJ)

            \footnotesize

            \item Cause retirement? \\ \vspace{1mm}
            \tiny (Schleife 2006, Labour; Cavapozzi et al 2013, AE; Roger et al 2006, EJ; Friedburg 2003, ILRR)

            \footnotesize

            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
            \tiny (Willis and Hudomiet 2021, Labour)

            \footnotesize

            \item Computerization in white collar jobs leads to decrease in employment for the same jobs that implemented the technology \\ \vspace{1mm}
            \tiny (Dillender & Forsythe 2022, NBER)
        \end{itemize}

        \column{0.4\textwidth}
        \centering

        \color{blue}
        \underline{ Contribution }

        \vspace{1mm}

            \begin{itemize}
            \color{blue}
                \item While doctors are white collar, their labor market incentives differ from many other jobs
                \item Population of physicians - possible shortages / access to care issues
            \end{itemize}

\end{columns}
\end{frame}

\begin{frame}{Literature}
\small
\begin{columns}
\setlength{\tabcolsep}{-5pt}
        \column{0.4\textwidth}
        \centering
        \underline{ EHR $\rightarrow$ Healthcare }
        \vspace{-1mm}
        \begin{itemize}
            \item Small improvements for severe health conditions \\ \vspace{1mm}
            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND)

            \footnotesize

            \item No decrease in cost \\ \vspace{1mm}
            \tiny (Agha 2014, JHE; Dranove et al 2019, AEJ)

            \footnotesize

            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
            \tiny (Hitt $\&$ Tambe 2016, ILRR; Butler $\&$ Johnson 2016, IJHEM; Meyerhoefer et al 2016, ILRR)
        \end{itemize}

        \column{0.4\textwidth}
        \centering
        \color{blue}
        \underline{ Contribution }
        \vspace{1mm}
        \begin{itemize}
        \color{blue}
            \item No empirical work on EHR/physician effects
            \item Speaks to the puzzling results
            \item Time period 
        \end{itemize}

\end{columns}

\end{frame}

\section{Data}

\begin{frame}{Data}
    \underline{CMS Shared Patient Data (2009-2015)}
    \begin{itemize}
        \item Hospital-physician pairs: bill for same patient under Medicare
        \item Physician (PCP, hospitalist, internist) works in the hospital
    \end{itemize}

@@

    \underline{Medicare Data on Provider Practice and Specialty (2009-2017)}
    \begin{itemize}
        \item Physician-level number of unique patients, primary locations of practice, and total claims 
    \end{itemize}

@@

    \underline{EHR Use:}
    \begin{itemize}
        \item AHA Survey Question
    \end{itemize}

@@

    \underline{Physician Compare}: Age of physician

    $\rightarrow$ Final data is at the physician level from 2009-2017, where only physicians who are exposed by 2015 are included 
\end{frame}

\begin{frame}{Variables}

Labor Market Outcomes:
\begin{itemize}
    \item Retirement
    \begin{itemize}
        \item Physicians stop seeing patients $\rightarrow$ drop out of data
    \end{itemize}

    \item Work Setting:
    \begin{itemize}
        \item Percentage of patients seen in an office setting
        \item Indicator for seeing any patients in an office setting
        \item Indicator for changing zip codes
    \end{itemize}

    \item Productivity:
    \begin{itemize}
        \item Number of patients
        \item Ratio of claims per patient
    \end{itemize}
\end{itemize}

EHR Use:
\vspace{-2mm}
\begin{itemize}
    \item Indicator for any exposure to EHR at related hospitals
\end{itemize}

\end{frame}

\section{Analysis}

\begin{frame}{Event Study}

\begin{itemize}
    \item Since I have differential timing of exposure, I use Callaway and Sant'Anna (2021) average group time treatment effects
    \vspace{3mm}
    $$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1]$$
    \vspace{3mm}
    \item Aggregate these over groups to get a familiar event study plot
\end{itemize}

\end{frame}

\begin{frame}{Assumptions}
\begin{itemize}
    \item No reversal of treatment
    \item No anticipation of treatment (can be relaxed)
    \item Parallel trends
    \item Institutional assumption: physicians are constrained to use to technology implemented in their hospital
\end{itemize}

\end{frame}

\section{Results}

\begin{frame}{Retirement}
\begin{figure}[ht]
    \centering
    \includegraphics[scale=.4]{}
\end{figure}
\end{frame}

\begin{frame}{Number of Patients Seen in Office}
\begin{figure}[ht]
    \centering
    \includegraphics[scale=.4]{}
\end{figure}
\end{frame}

\begin{frame}{Indicator for working in office}
\begin{figure}[ht]
    \centering
    \includegraphics[scale=.4]{}
\end{figure}
\end{frame}

\begin{frame}{Change Zip}
\begin{figure}[ht]
    \centering
    \includegraphics[scale=.4]{}
\end{figure}
\end{frame}

\begin{frame}{Patient Count}
\begin{figure}[ht]
    \centering
    \includegraphics[scale=.4]{}
\end{figure}
\end{frame}

\begin{frame}{Claims per Patient}
\begin{figure}[ht]
    \centering
    \includegraphics[scale=.4]{}
\end{figure}
\end{frame}

\begin{frame}{Summary}
\begin{itemize}
    \item Evidence that physicians are changing their behavior because of this new technology, specifically through retirement or job switching
    \item Little evidence of work setting changes
    \item For physicians who endure through the implementation of the technology, claims per patient goes down.

@@

\section{Continuing Work}

\begin{frame}{Plans for this paper}
    \begin{itemize}
        \item Analysis on retirement decision of physicians: requires a better understanding of why physicians would drop out of this Medicare data
        \item See if there are any specific years driving these results
        \item Endogenous treatment? (Bandwidth, state privacy laws as IV)
    \end{itemize}
\end{frame}

\begin{frame}[plain]{}
\centering
    Thank you!
\end{frame}

>>>>

\begin{frame}{Big Picture}
How does technology impact employment?

    \begin{itemize}
        \item Displacement?
        \vspace{2mm}
        \item Productivity?
    \end{itemize}
    \vspace{3mm}

Investigated in different ways

    \begin{itemize}
        \item Computerization on white collar jobs
        \vspace{2mm}
        \item Machines on factory workers
    \end{itemize}

@@

    \begin{alertblock}{$\rightarrow$ This paper considers this question in the context of healthcare and physician labor markets}\end{alertblock}
\end{frame}

\begin{frame}[fragile]{What is an Electronic Health Record (EHR)?}
``An electronic health record is the systematized collection of patient and population electronically stored health information in a digital format"

\vspace{5mm}
\centering
\includegraphics[scale=.35]{Objects/EHR_picture.PNG}
\end{frame}

@@

Significant policy push for EHR implementation: HITECH Act, 2008 provided financial incentive for hospitals to implement EHRs
\begin{itemize}
    \item \textcolor{blue}{The percentage of hospitals with basic EHR capability rose from 9$\%$ in 2008 to 84$\%$ in 2015.} \scriptsize (Health IT Dashboard)
\end{itemize}
\end{frame}

@@

    \includegraphics[scale=.3]{graphics/News Clip2.PNG}
\end{center}
\end{frame}

\begin{frame}{This Paper}
$\rightarrow$ Did EHR implementation in hospitals affect physician labor market decisions?

\vspace{5mm}
\centering
\includegraphics[scale=.45]{Objects/EHR_FlowChart_General.PNG}
\end{frame}

\section{Analysis and Data}

\begin{frame}{Difference in Difference Analysis}
Treatment: Exposure to EHR

@@

Since I have differential timing of exposure, I use Callaway and Sant'Anna (2021) average group time treatment effects

$$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1]$$

@@

\begin{itemize}
\item Not-yet-treated units as control group
    \item Aggregate these over groups to get a familiar event study plot
\end{itemize}
\end{frame}

\begin{frame}{Data}
Panel of hospitalists from 2009-2017 with information on whether the hospital uses an EHR and various labor market outcomes

\vspace{4mm}
\pause
\begin{itemize}
    \item EHR information ends in 2015, so I limit to only those who were exposed by 2015
\end{itemize}
\end{frame}

\begin{frame}{Assumptions}
\begin{enumerate}
    \item No reversal of treatment
    \vspace{3mm}
    \item No anticipation of treatment (can be relaxed)
    \vspace{3mm}
    \item Parallel trends
\end{enumerate}
\end{frame}

\section{Retirement}

\begin{frame}{Retirement}
    \centering
    \includegraphics[scale=.5]{Objects/EHR_FlowChart_Retire.PNG}
\end{frame}

\begin{frame}{Retirement Variable}
Indicator variable equal to 1 if the physician drops out of the data 
\vspace{3mm}
\begin{itemize}
    \item No longer seeing patients
\end{itemize}
\end{frame}

\begin{frame}{Physician Incentives}
Decision to retire is different than we usually think of
\vspace{3mm}
    \begin{itemize}
        \item Retirement is thought of as a function of pensions, and sometimes depends on shocks such as health or job loss
        \tiny (Hall and Johnson 1980, ILRR), (Gordon and Blinder 1980, J Public Econ), (Hamermesh 1985, JOLE), $\cdot$

@@

        \normalsize \item On average, physicians plan to retire at age 60 but don't actually retire until age 69 \tiny(Collier 2017)
    \end{itemize}

\vspace{3mm}
\pause

\normalsize Unique set of incentives for which a shock to the work setting \textit{may} lead to retirement more-so than other occupations, for which studies yield mixed results  \tiny (Schleife 2006, Labour; Cavapozzi et al 2013, AE; Roger et al 2006, EJ; Friedburg 2003, ILRR)

\vspace{3mm}
\pause

\normalsize Related studies find that well educated workers exited at higher rates due to computerization \tiny (Willis and Hudomiet 2021, Labour), (Dillender & Forsythe 2022, NBER)
\end{frame}

\begin{frame}{Leaving Clinical Setting}
    \begin{itemize}
        \item Retirement is loosely defined
        \vspace{5mm}
        \item Any age physician may decide to leave a clinical setting
    \end{itemize}
\end{frame}

\begin{frame}{Retirement}
\begin{figure}[ht]
    \centering
    \includegraphics[scale=.35]{Objects/Presentation_retire_all.pdf}
\end{figure}
$\rightarrow$ EHR exposure led to a .003 (.0045) percentage point increase in the likelihood of retirement in the first (second) year after exposure (10\% and 15\% increase relative to the mean)
\end{frame}

\begin{frame}{Retirement: Senior Physicians}
\begin{figure}[ht]
    \centering
    \includegraphics[scale=.35]{Objects/Presentation_retire_old.pdf}
\end{figure}
$\rightarrow$ 25\% increase relative to the mean in the first year, but 0 in the second\\
$\rightarrow$ What does this mean for younger physicians?
\end{frame}

\section{Job Setting}

\begin{frame}{Changing Job Setting}
    \centering
    \includegraphics[scale=.5]{Objects/EHR_FlowChart_JobSwitch.PNG}
\end{frame}

\begin{frame}{Job Switching Variables}
    \begin{enumerate}
        \item Fraction of Patients Seen in Office 
        \vspace{3mm}
        \item Indicator for working in office or not
    \end{enumerate}
\end{frame}

\begin{frame}{Physician Incentives}
Most physicians will not be induced to retire because of EHRs
    \begin{itemize}
        \item Another way to avoid (or be exposed to) new technology is to switch location of practice
        \begin{itemize}
            \vspace{3mm}
            \item Literature shows job switching is closely tied to job satisfaction \tiny (Aklerlof, Rose and Yellen 1988, Brookings), (Chadi and Hetschko 2017, JEMS)
        \end{itemize}
        \vspace{6mm}
        \pause
        \item The cost associated with switching depends on the strength of the tie with the hospital
    \end{itemize}
\end{frame}

\begin{frame}{Fraction of Patients Seen in Office}
\begin{figure}[ht]
    \centering
    \includegraphics[scale=.35]{Objects/Presentation_fracoffice_all.pdf}
\end{figure}
$\rightarrow$ No effect!
\end{frame}

\begin{frame}{Indicator for working in office}
\begin{figure}[ht]
    \centering
    \includegraphics[scale=.35]{Objects/Presentation_office_all.pdf}
\end{figure}
$\rightarrow$ 4.6\% increase relative to the mean\\
$\rightarrow$ Driven by young physicians
\end{frame}

\section{Productivity}

\begin{frame}{Productivity}
    \centering
    \includegraphics[scale=.45]{Objects/EHR_FlowChart_Productivity.PNG}
\end{frame}

\begin{frame}{Productivity Variables}
    \begin{enumerate}
        \item Patient Count
        \vspace{6mm}
        \item Claims per patient
    \end{enumerate}
\end{frame}

\begin{frame}{Physician Incentives}
Past research on information technology $\rightarrow$ productivity typically indicates increase \tiny (Bartel, Ichniowski and Shaw 2007, QJE)
\vspace{3mm}
\begin{itemize}
    \normalsize \item Sometimes depends on managerial style \tiny(Garicano and Heaton 2010, JLE)
\end{itemize}

\vspace{6mm}

\normalsize We can see how productivity is affected for physicians who don't change behavior as a result of EHRs
\vspace{3mm}
    \begin{itemize}
        \item Expected effect is ambiguous
        \vspace{3mm}
        \item Claim Count per patient is a matter of physician incentives
    \end{itemize}
\end{frame}

\begin{frame}{Patient Count}
\begin{figure}[ht]
    \centering
    \includegraphics[scale=.35]{Objects/Presentation_patients_all.pdf}
\end{figure}
$\rightarrow$ 5-10\% increase relative to the mean\\
$\rightarrow$ Driven by demand? 
\end{frame}

\begin{frame}{Patient Count Broken Down by Group}
\centering
    \includegraphics[scale=.25]{Objects/patient_group.pdf}
\end{frame}

\begin{frame}{Patient Count: Senior Physicians}
\begin{figure}[ht]
    \centering
    \includegraphics[scale=.4]{Objects/Presentation_patients_old.pdf}
\end{figure}
\end{frame}

\begin{frame}{Claims per Patient}
\begin{figure}[ht]
    \centering
    \includegraphics[scale=.35]{Objects/Presentation_claimperpatient_all.pdf}
\end{figure}
$\rightarrow$ No significant change, but a downward trend is possible
\end{frame}

\section{Conclusion}

\begin{frame}{Summary}
\begin{itemize}
    \item Evidence that physicians are changing their behavior because of this new technology, specifically through retirement or job switching
    \vspace{3mm}
    \item Little evidence of work setting changes
    \vspace{3mm}
    \item For physicians who endure through the implementation of the technology, number of patients goes up but not conclusive evidence that claims per patient goes down

@@

\begin{frame}{Robustness Checks}
    \begin{enumerate}
        \item Endogeneity of treatment
        \item Data assistants
        \item Limiting Years and Using Never-treated as control
        \item Anticipation
    \end{enumerate}
\end{frame}

\begin{frame}[plain]{}
\centering
    Thank you! \\
    \vspace{5mm}
    Email: hkagele@emory.edu
\end{frame}
